---
title: "Assessing Pesticide Contamination in Illinois Water Supply and the Impact of Agricultural Practices"
author: Carson Edwards
output: html_notebook
---

1) Load in libraries

```{r}
library(purrr)
library(readr)
library(corrplot)
library(forecast)
library(tidyr)
library(randomForest)
library(xgboost)
library(changepoint)
library(urbnmapr)
library(ggplot2)
library(dplyr)
library(data.table)
```

##############################################################################################################################

                                                TRI PROGRAM DATASET ANALYSIS SECTION 

##############################################################################################################################

2) Load and format TRI program data

```{r}
folder_path <- "C:/Users/carso/Downloads/il_chemicals/"
file_list <- list.files(path = folder_path, pattern = "\\d{4}_il\\.csv$", full.names = TRUE)

combined_data <- map_df(file_list, ~ read_csv(.x, col_types = cols(.default = "c")))

# Clean up column names
cleaned_colnames <- colnames(combined_data) %>%
  gsub("^\\d+\\.\\s*", "", .) %>%  
  gsub("\\s+", "_", .) %>%        
  tolower()                       
colnames(combined_data) <- cleaned_colnames

# Mutate total releases to numeric, convert yes/no to logical, and as numeric
combined_data <- combined_data %>%
  mutate(
    total_releases = as.numeric(total_releases),
    across(where(~ is.character(.) & all(. %in% c("YES", "NO"))), ~ . == "YES"),
    across(where(is.logical), as.numeric)
  )

# Get sums of county, chemical, and year combinations
chemical_by_county <- combined_data %>%
  group_by(county, chemical, year) %>%
  summarise(total_releases = sum(total_releases, na.rm = TRUE))

# Load in our county map data
illinois_counties <- get_urbn_map("counties", sf = FALSE) %>%
  filter(state_abbv == "IL")

head(combined_data)
head(chemical_by_county)
```

3) Correlation modeling

```{r}
# Selecting our numeric data, we have already converted true/false to 1/0
numeric_data <- combined_data %>%
  select_if(is.numeric)
correlation_matrix <- cor(numeric_data, use = "complete.obs")

correlation_matrix

colnames(correlation_matrix) <- gsub("_", " ", colnames(correlation_matrix))
rownames(correlation_matrix) <- gsub("_", " ", rownames(correlation_matrix))

# Paste into console to see a readable plot in plot window
corrplot(correlation_matrix, 
         method = "color", 
         tl.cex = 1.5, 
         tl.col = "black",  
         addCoef.col = "black")  
```

4) Random forest model for feature importance

```{r}
# Only selecting fields of interest
model_data <- combined_data %>%
  select(total_releases, year, federal_facility, elemental_metal_included, clean_air_act_chemical, metal, carcinogen, pbt, pfas) 

set.seed(123)  
rf_model <- randomForest(total_releases ~ ., data = model_data, importance = TRUE, ntree = 500)

print(rf_model)
importance(rf_model)
varImpPlot(rf_model, 
           main = "Random Forest Classifier for TRI Program Variable Importance")
```


5) Change point analysis for TRI program data

```{r}
get_change_points <- function(df) {
  df$year <- as.numeric(df$year)
  df <- df[order(df$year), ]
  if (any(is.na(df$total_releases)) || !is.numeric(df$total_releases)) {
    return(NULL)
  }
  
  releases <- as.numeric(df$total_releases)  
  max_Q <- length(releases) - 1

  # The max change points to search for. Taking either length of county chemical combinations or total number of rows. Want less change points rather than more
  Q <- min(3, max_Q)
  if (Q < 1) {
    return(NULL)
  }
  
  # Was getting too many warnings with the output
  suppressWarnings({
    cpt <- cpt.mean(releases, method = "BinSeg", penalty = "BIC", Q = Q)
  })

  return(cpts(cpt))
}

change_points <- list()

# Only want combinations with at least 3 reported years
valid_data <- chemical_by_county %>%
  group_by(county, chemical) %>%
  filter(n() >= 3) %>%
  ungroup()

# Need unique pairs 
unique_combinations <- unique(valid_data[, c("county", "chemical")])

# For each unique combination of county and chemical, take the subset of data and run get change points function
for (i in 1:nrow(unique_combinations)) {
  county1 <- unique_combinations$county[i]
  chemical1 <- unique_combinations$chemical[i]
  subset_data <- subset(valid_data, county == county1 & chemical == chemical1)
  
  result <- get_change_points(subset_data)

  if (!is.null(result)) {
    change_points[[paste(county1, chemical1, sep = "_")]] <- result
  } else {
    change_points[[paste(county1, chemical1, sep = "_")]] <- "No valid data or change points"
  }
}
```

6) Functions for plotting change points and getting year counts

```{r}
plot_change_points <- function(df, change_points) {
  df$year <- as.numeric(df$year)
  
  # Order by year
  df <- df[order(df$year), ]
  
  # Create change point plot of total releases
  change_point_plot <- ggplot(df, aes(x = year, y = total_releases)) +
    geom_line() +
    geom_point() +
    labs(title = paste(df$county[1], df$chemical[1], "Change Point Detection"),
         x = "Year", y = "Total Releases") +
    theme_minimal()
  
  # Plotting change points using a vertical red line 
  if (!is.null(change_points) && length(change_points) > 0) {
    change_point_plot <- change_point_plot + geom_vline(xintercept = df$year[change_points], color = "red", linetype = "dashed")
  }
  print(change_point_plot)
}

analyze_change_points <- function(chemical_name, chemical_by_county, change_points) {
  unique_counties <- unique(chemical_by_county$county[chemical_by_county$chemical == chemical_name])
  change_point_years <- c()
  
  # For each county chemical combination, get the change point counts 
  for (county1 in unique_counties) {
    subset_data <- subset(chemical_by_county, county == county1 & chemical == chemical_name)
    change_point_result <- change_points[[paste(county1, chemical_name, sep = "_")]]
    
    if (!is.null(change_point_result)) {
      plot_change_points(subset_data, change_point_result)
      change_point_years <- c(change_point_years, change_point_result)  # Collect change point years
    }
  }
  
  change_point_year_counts <- table(change_point_years)
  
  # Plot the totals
  barplot(change_point_year_counts, 
          main = paste("Frequency of Change Point Years for", chemical_name), 
          xlab = "Year", 
          ylab = "Frequency", 
          col = "skyblue", 
          las = 2, 
          names.arg = as.character(as.numeric(names(change_point_year_counts)) + 2000)) # Have years display correctly

}
```

Ammonia example

```{r}
analyze_change_points("Ammonia", chemical_by_county, change_points)
```

Benzene example 

```{r}
analyze_change_points("Benzene", chemical_by_county, change_points)
```

Change point detection by year

```{r}
change_points <- list()

# Group by year and chemical this time
valid_data <- chemical_by_county %>%
  group_by(year, chemical) %>%
  summarise(total_releases = sum(total_releases, na.rm = TRUE)) %>%
  filter(n() >= 3) %>%
  ungroup()

# Only unique chemicals since no counties
unique_chemicals <- unique(valid_data$chemical)

# For each unique chemical, take the subset of data and run get change points function
for (i in 1:length(unique_chemicals)) {
  chemical1 <- unique_chemicals[i]
  subset_data <- subset(valid_data, chemical == chemical1)
  
  result <- get_change_points(subset_data)

  if (!is.null(result)) {
    change_points[[chemical1]] <- result
  } else {
    change_points[[chemical1]] <- "No valid data or change points"
  }
}

plot_change_points <- function(df, change_points) {
  df$year <- as.numeric(df$year)
  
  # Order by year
  df <- df[order(df$year), ]
   
  # Create change point plot of total releases
  change_point_plot <- ggplot(df, aes(x = year, y = total_releases)) +
    geom_line() +
    geom_point() +
    labs(title = paste(df$chemical[1], "Change Point Detection"),
         x = "Year", y = "Total Releases") +
    theme_minimal()
  
  # Plotting change points using a vertical red line 
  if (!is.null(change_points) && length(change_points) > 0) {
    change_point_plot <- change_point_plot + geom_vline(xintercept = df$year[change_points], color = "red", linetype = "dashed")
  }
  print(change_point_plot)
}

analyze_change_points <- function(chemical_name, chemical_by_county, change_points) {
  subset_data <- subset(chemical_by_county, chemical == chemical_name)  # Only chemical for this subset
  change_point_result <- change_points[[chemical_name]]
  
  if (!is.null(change_point_result)) {
    plot_change_points(subset_data, change_point_result)
    change_point_years <- change_point_result  # Collect change point years
  } else {
    change_point_years <- c()
  }
}
```

Ammonia change points statewide

```{r}
analyze_change_points("Ammonia", valid_data, change_points)
```

Benzene change points statewide

```{r}
analyze_change_points("Benzene", valid_data, change_points)
```

Atrazine change points statewide

```{r}
analyze_change_points("Atrazine", valid_data, change_points)
```

2,4-D change points statewide

```{r}
analyze_change_points("2,4-D", valid_data, change_points)
```


7) Geospatial analysis for TRI program data

Average yearly increase function

```{r}
plot_avg_yearly_increase <- function(chemical_name, chemical_by_county, illinois_counties) {
  
  # Filter by chemical of interest, group by county, arrange by year, calculate yearly change, take the mean and drop groups
  avg_increase_data <- chemical_by_county %>%
    filter(chemical == chemical_name) %>%
    group_by(county) %>%
    arrange(year) %>%
    mutate(yearly_change = total_releases - lag(total_releases)) %>%
    summarize(avg_yearly_increase = mean(yearly_change, na.rm = TRUE), .groups = "drop") %>%
    rename(county_name = county)
  
  # Convert county names to regular format ex "Jo Daviess County"
  avg_increase_data$county_name <- paste(tools::toTitleCase(tolower(avg_increase_data$county_name)), "County")
  
  # Set up counties data with average increase data by county name
  spatial_data <- left_join(illinois_counties, avg_increase_data, by = "county_name")
  
  # Plot using Albers projection to minimize warping, remove axis labels, and apply nice color scheme
  plot <- ggplot() +
    geom_polygon(data = spatial_data,
                 mapping = aes(x = long, y = lat, group = group, fill = avg_yearly_increase),
                 color = "#ffffff", size = 0.25) +
    coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
    scale_fill_gradient2(midpoint = 0, low = "blue", mid = "lightblue", high = "orange", 
                     na.value = "grey", name = "Average Yearly Increase in lbs") +
    labs(title = paste("Average Yearly Increase in Total Releases of", chemical_name)) +
    theme_minimal() +
    theme(axis.title = element_blank(),
          axis.text = element_blank(),
          axis.ticks = element_blank()) 
  
  print(plot)
}
```

Ammonia average yearly increase (Jo Daviess county of interest)

```{r}
plot_avg_yearly_increase("Ammonia", chemical_by_county, illinois_counties)
```

Benzene average yearly increase (Benzene generally decreasing)

```{r}
plot_avg_yearly_increase("Benzene", chemical_by_county, illinois_counties)
```

2,4-D average yearly increase (Cook county of interest)

```{r}
plot_avg_yearly_increase("2,4-D", chemical_by_county, illinois_counties)
```

Mercury average yearly increase 

```{r}
plot_avg_yearly_increase("Mercury", chemical_by_county, illinois_counties)
```

Hydrazine average yearly increase

```{r}
plot_avg_yearly_increase("Hydrazine", chemical_by_county, illinois_counties)
```

Atrazine average yearly increase

```{r}
plot_avg_yearly_increase("Atrazine", chemical_by_county, illinois_counties)
```
Acetochlor shows no reported releases

```{r}
plot_avg_yearly_increase("Acetochlor", chemical_by_county, illinois_counties)
```

##############################################################################################################################

                                             USGS WATER QUALITY DATASET ANALYSIS SECTION 

##############################################################################################################################

1) Load in and format the USGS data

```{r}
chemical_data <- fread("C:/Users/carso/Downloads/waterdata/waterdata.csv", sep = "\t", fill = TRUE, quote = "")

# Format date field
chemical_data$sample_dt <- as.Date(chemical_data$sample_dt, format="%Y-%m-%d")
chemical_data <- chemical_data %>% drop_na(result_va)
```

Examine data

```{r}
head(chemical_data)
summary(chemical_data)
```

2) Load in and format parameter data 

```{r}
param_data <- read.delim("C:/Users/carso/Downloads/waterdata/param_dictionary.txt", sep = "\t", header = FALSE, stringsAsFactors = FALSE, col.names = "field")

# Format param data so we can match it. Don't need field
param_data <- param_data %>%
  mutate(
    parm_cd = as.numeric(sub(" - .*", "", field)),
    description = sub("^[0-9]+ - ", "", field) 
  ) %>%
  select(-field)   

```

3) Function for filtering by chemicals in the parameter codes and using a SARIMA model to forecast

```{r}
forecast_chemical <- function(chemical_name, start_date, param_data, chemical_data, forecast_horizon = 36) {
  
  # Get data for chemical of interest using grepl
  data <- param_data %>%
    filter(grepl(chemical_name, description, ignore.case = TRUE))
  
  # Get parameter codes that match that chemical and select data after start date
  filtered_chemical_data_after_start <- chemical_data %>%
    filter(parm_cd %in% data$parm_cd, sample_dt > as.Date(start_date)) %>%
    select(site_no, sample_dt, sample_end_dt, result_va, result_lab_cm_tx) %>%
    mutate(year_month = format(sample_dt, "%Y-%m"))
  
  # Get outlier threshold
  outlier_threshold <- quantile(filtered_chemical_data_after_start$result_va, 0.99, na.rm = TRUE)
  
  # Remove outliers above the threshold
  filtered_chemical_data_no_outliers <- filtered_chemical_data_after_start %>%
    filter(result_va <= outlier_threshold)
  
  # Get data by year month and get mean
  data_by_month <- filtered_chemical_data_no_outliers %>%
    group_by(year_month) %>%
    summarise(result_va = mean(result_va, na.rm = TRUE)) %>%
    ungroup() %>%
    arrange(year_month)
  
  # Convert year month to date
  data_by_month <- data_by_month %>%
    mutate(year_month = as.Date(paste0(year_month, "-01")))
  
  # Create time series using the data
  result_va_ts_monthly <- ts(
    data_by_month$result_va, 
    start = c(
      as.numeric(format(min(data_by_month$year_month), "%Y")), 
      as.numeric(format(min(data_by_month$year_month), "%m"))
    ), 
    frequency = 12
  )
  
  # Fit SARIMA model
  sarima_model <- auto.arima(result_va_ts_monthly, seasonal = TRUE)
  
  # Get forecast
  forecast_monthly <- forecast(sarima_model, h = forecast_horizon)
  
  # Plot forecast
  plot(
    forecast_monthly, 
    main = paste("Forecast of Monthly Average result_va for", chemical_name), 
    xlab = "Year", 
    ylab = "Average Amount in Water Supply"
  )
  
  # Extract forecasted values
  forecast_values <- as.data.frame(forecast_monthly)$`Point Forecast`
  
  # Calculate the difference between the first and last forecasted values
  forecast_increase <- forecast_values[length(forecast_values)] - forecast_values[1]
  
  # Print increase or decrease
  if (forecast_increase > 0) {
    cat("The forecast indicates an increase of", forecast_increase, "over the next", forecast_horizon, "months.\n")
  } else if (forecast_increase < 0) {
    cat("The forecast indicates a decrease of", abs(forecast_increase), "over the next", forecast_horizon, "months.\n")
  } else {
    cat("The forecast indicates no significant change over the forecast period.")
  }
  
  # Print summary
  print(summary(sarima_model))
  
  # Return the forecast and model summary
  return(list(
    model = sarima_model,
    forecast = forecast_monthly
  ))
  
}

```

2,4-D SARIMA forecast post 2014 since the EPA approved use in 2014. It picks up a decrease but when reading the chart you can see its just the top end of the season its starting at so note this as possible increase

```{r}
result <- forecast_chemical(
  chemical_name = "2,4-D",
  start_date = "2014-01-01",
  param_data = param_data,
  chemical_data = chemical_data,
  forecast_horizon = 36
)
result

```

Ammonia SARIMA forecast. Slight increase detected though it starts at the bottom of the season. Peaks are higher for years post 2020.

```{r}
result <- forecast_chemical(
  chemical_name = "Ammonia",
  start_date = "2000-01-01",
  param_data = param_data,
  chemical_data = chemical_data,
  forecast_horizon = 36
)
result
```

Atrazine SARIMA forecast. It picks up a decrease but when reading the chart you can see its just the top end of the season its starting at. Note this as possible increase because the bottom line of the prediction looks like it is increasing.
```{r}
result <- forecast_chemical(
  chemical_name = "Atrazine",
  start_date = "2000-01-01",
  param_data = param_data,
  chemical_data = chemical_data,
  forecast_horizon = 60
)
result
```

```{r}
result <- forecast_chemical(
  chemical_name = "Atrazine",
  start_date = "2012-01-01",
  param_data = param_data,
  chemical_data = chemical_data,
  forecast_horizon = 60
)
result
```

```{r}
result <- forecast_chemical(
  chemical_name = "Atrazine",
  start_date = "2012-01-01",
  param_data = param_data,
  chemical_data = chemical_data,
  forecast_horizon = 36
)
result
```

Benzene SARIMA forecast. Not enough data to predict anything with the model.

```{r}
result <- forecast_chemical(
  chemical_name = "Benzene",
  start_date = "2000-01-01",
  param_data = param_data,
  chemical_data = chemical_data,
  forecast_horizon = 36
)
result
```

Metolachlor SARIMA forecast. Found post 2012 was where most measurements happen. Increase detected

```{r}
result <- forecast_chemical(
  chemical_name = "Metolachlor",
  start_date = "2012-01-01",
  param_data = param_data,
  chemical_data = chemical_data,
  forecast_horizon = 60
)
result
```

```{r}
result <- forecast_chemical(
  chemical_name = "Metolachlor",
  start_date = "2012-01-01",
  param_data = param_data,
  chemical_data = chemical_data,
  forecast_horizon = 36
)
result
```

Acetochlor SARIMA forecast. Increase detected

```{r}
result <- forecast_chemical(
  chemical_name = "Acetochlor",
  start_date = "2013-01-01",
  param_data = param_data,
  chemical_data = chemical_data,
  forecast_horizon = 36
)
result
```

Diazinon SARIMA forecast. Data has inconsistent drops and peaks, no forecast

```{r}
result <- forecast_chemical(
  chemical_name = "Diazinon",
  start_date = "2000-01-01",
  param_data = param_data,
  chemical_data = chemical_data,
  forecast_horizon = 36
)
result
```

Azoxystrobin SARIMA forecast. Increase detected

```{r}
result <- forecast_chemical(
  chemical_name = "Azoxystrobin",
  start_date = "2000-01-01",
  param_data = param_data,
  chemical_data = chemical_data,
  forecast_horizon = 36
)
result
```



